---
title: "BA Group Project"
author: "GROUP_7"
date: "2024-11-29"
output: html_document
---

## Loading Required Libraries

```{r}
# Load necessary libraries
library(dplyr)
library(ggplot2)
library(caret)
library(tidyverse)
library(ROCR)
library(randomForest)
library(rpart)
library(rpart.plot)
library(rattle)

```

Loading and Exploring the Data

```{r}
# Load the training data
churn_data <- read.csv("~/Downloads/Churn_Train.csv")

# Explore the data
dim(churn_data)
str(churn_data)
summary(churn_data)

# Check for missing values
colSums(is.na(churn_data))

```

Visualizing Data

```{r}
ggplot(churn_data, aes(x = churn, fill = churn)) +
  geom_bar(color = "black") +
  scale_fill_manual(values = c("yes" = "purple", "no" = "lavender")) +
  ggtitle("Churn Distribution") +
  xlab("Churn Status") +
  ylab("Count") +
  theme_minimal()
```

Preprocessing the Data
Handle Categorical Variables

```{r}
# Convert categorical variables to factors
churn_data$churn <- as.factor(churn_data$churn)
churn_data$international_plan <- as.factor(churn_data$international_plan)
churn_data$voice_mail_plan <- as.factor(churn_data$voice_mail_plan)
```
Impute Missing Values

```{r}
# Define a function to calculate mode
get_mode <- function(x) {
  uniq_vals <- unique(x)
  uniq_vals[which.max(tabulate(match(x, uniq_vals)))]
}

# Impute missing values in categorical variables using mode
factor_columns <- sapply(churn_data, is.factor)
churn_data[factor_columns] <- lapply(churn_data[factor_columns], function(col) {
  if (any(is.na(col))) {
    col[is.na(col)] <- get_mode(col)
  }
  return(col)
})

# Impute missing values in numeric variables using mean
numeric_columns <- sapply(churn_data, is.numeric)
churn_data[numeric_columns] <- lapply(churn_data[numeric_columns], function(col) {
  ifelse(is.na(col), mean(col, na.rm = TRUE), col)
})
```

Split the Data

```{r}
# Split the data into training and testing sets
set.seed(123)
train_index <- createDataPartition(churn_data$churn, p = 0.8, list = FALSE)
train_data <- churn_data[train_index, ]
test_data <- churn_data[-train_index, ]
```

Modeling
Logistic Regression 
# Train a logistic regression model

```{r}
set.seed(1)
logistic_model <- glm(churn ~ ., data = train_data, family = binomial)

# View model summary
summary(logistic_model)
```

#Random Forest

# Train a Random Forest model
```{r}
set.seed(2)
random_forest_model <- randomForest(churn ~ ., data = train_data, ntree = 100, mtry = 3)

# Variable importance plot
varImpPlot(random_forest_model, main = "Variable Importance")
```
# Decision Tree
```{r}
# Train a Decision Tree model
set.seed(3)
dt_model <- rpart(churn ~ ., data = train_data, method = "class", control = rpart.control(minsplit = 150))

# Visualize the Decision Tree
fancyRpartPlot(dt_model)
```

Model Evaluation
Logistic Regression

```{r}
# Predict probabilities on the test set
logistic_preds <- predict(logistic_model, newdata = test_data, type = "response")

# Convert probabilities to binary predictions
predicted_class <- ifelse(logistic_preds > 0.5, "yes", "no")

# Confusion matrix
confusionMatrix(as.factor(predicted_class), test_data$churn)

# ROC Curve
pred <- prediction(logistic_preds, test_data$churn)
perf <- performance(pred, "tpr", "fpr")
plot(perf, col = "blue", main = "ROC Curve")
abline(a = 0, b = 1, lty = 2)

# Calculate AUC
auc_perf <- performance(pred, "auc")
auc_value <- auc_perf@y.values[[1]]
print(paste("AUC:", auc_value))
```
Random Forest

```{r}
# Predict on the test set
rf_preds <- predict(random_forest_model, newdata = test_data, type = "response")

# Confusion matrix
confusionMatrix(as.factor(rf_preds), test_data$churn)

# ROC Curve
rf_pred_probs <- predict(random_forest_model, newdata = test_data, type = "prob")[, 2]
rf_pred <- prediction(rf_pred_probs, test_data$churn)
rf_perf <- performance(rf_pred, "tpr", "fpr")
plot(rf_perf, col = "green", main = "ROC Curve - Random Forest")
abline(a = 0, b = 1, lty = 2)

# Calculate AUC
rf_auc_perf <- performance(rf_pred, "auc")
rf_auc_value <- rf_auc_perf@y.values[[1]]
print(paste("AUC - Random Forest:", rf_auc_value))

```

Decision Tree
```{r}
# Predict on the test set
dt_preds <- predict(dt_model, newdata = test_data, type = "class")

# Confusion matrix
confusionMatrix(as.factor(dt_preds), test_data$churn)

# ROC Curve
dt_pred_probs <- predict(dt_model, newdata = test_data, type = "prob")[, 2]
dt_pred <- prediction(dt_pred_probs, test_data$churn)
dt_perf <- performance(dt_pred, "tpr", "fpr")
plot(dt_perf, col = "orange", main = "ROC Curve - Decision Tree")
abline(a = 0, b = 1, lty = 2)

# Calculate AUC
dt_auc_perf <- performance(dt_pred, "auc")
dt_auc_value <- dt_auc_perf@y.values[[1]]
print(paste("AUC - Decision Tree:", dt_auc_value))

```
Predict Churn for New Data
```{r}
# Load the new data
load("~/Downloads/Customers_To_Predict.RData")
```

View the First Few Rows of the Data
```{r}
head(Customers_To_Predict)
```

# Preprocess the new data (convert categorical variables to factors)
```{r}
Customers_To_Predict$international_plan <- as.factor(Customers_To_Predict$international_plan)
Customers_To_Predict$voice_mail_plan <- as.factor(Customers_To_Predict$voice_mail_plan)
```

# Impute missing values in numeric columns
```{r}
numeric_columns <- sapply(Customers_To_Predict, is.numeric)
Customers_To_Predict[numeric_columns] <- lapply(Customers_To_Predict[numeric_columns], function(col) {
  ifelse(is.na(col), mean(col, na.rm = TRUE), col)
})
```
# Predict churn probabilities
```{r}
Customers_To_Predict$Churn_Prob <- predict(random_forest_model, newdata = Customers_To_Predict, type = "prob")[, 2]
```
# Save predictions to a CSV file
```{r}
# Construct the path for the output CSV file
file_path <- "/Users/dimplefnu/Downloads/BA Group_7 /New_Customers_To_Predict.csv"

# Write the CSV file
write.csv(Customers_To_Predict, file_path, row.names = FALSE)
```

```{r}
# Define the threshold
threshold <- 0.5

# Add churn classification based on the threshold
Customers_To_Predict <- Customers_To_Predict %>%
  mutate(churn = ifelse(Churn_Prob > threshold, "yes", "no"))

# View the updated dataset
head(Customers_To_Predict[, c("Churn_Prob", "churn")])
```

```{r}
 write.csv(Customers_To_Predict,"/Users/dimplefnu/Downloads/BA Group_7 /Prediction Group_7.csv", row.names = FALSE)
```